Visualizing Independence Using Extended Association Plots

### 2005-08-19 trunk/vcd/paper/JCGS.Rnw
comments by DM included
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@442 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

explore independence structure in contingency tables by log-linear models and visually
by mosaic displays

independence in 2-way tables

association and mosaic displays

colors and color spaces
    because the three perceptual dimensions (hue, lightness, saturation)
    are not properly mapped to the three dimensions of the color space
    and hence are confounded \citep{vcd:Brewer:1999}

somewhere:
mention Augsburg stuff, visualization of log-linear models via mosaics
\citep{vcd:Theus+Lauer:1999,vcd:Hofmann:2003,vcd:Hofmann:2001}.

@@

@@

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced first in statistics
textbooks if they address the independence problem in 2-way tables
\citep[see e.g.,][]{vcd:Agresti:2002}.

\begin{table}[t!]
\begin{center}
\caption{\label{tab:art} Treatment and improvement among $59$ patients
with rheumatoid arthritis.}

@@

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<classic_plots,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, margins = c(2.5, 4, 2.5, 3))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
assoc(art, newpage = FALSE, margins = c(5, 2, 5, 4))
popViewport(2)
@
\caption{Classic mosaic and association plot for the arthritis data.}
\label{fig:classic}
\end{center}
\end{figure}

@@

\section{Extension to Conditional Independence}

@@

\section{Conclusions}

Various strategies for constructing residual-based shadings for visualizing
(conditional) independence in contingency tables via mosaic and association plots
are discussed. The shading
of \cite{vcd:Friendly:1994} is extended in two directions: the use of perceptually
uniform HCL colors and the combination of visualization and significance testing.
To achieve the former, a general guideline for constructing diverging palettes
in HCL space is introduced. The advantages of using this HCL shading scheme
instead of an HSV scheme are that the colors from this perceptually-based color space
provide uniform saturations over different hues and that the colorfulness in this
shading can be controlled independently from the other two dimensions.

To combine visualization and significance testing, two approaches are presented:
The first approach always uses a fully colored palette but relies on
data-driven cut offs such that the presence of color is equivalent to significance of the
associated maximum test. The second approach uses pre-defined cut offs (such as $2$ and $4$) but
codes the result of the significance test by using full colors only for significant results and
the same type of palette with a reduced amount of color for non-significant results.

Both results can not only be applied to the simple independence problem in 2-way
contingency tables, but also to arbitrary inpendence models fitted via log-linear models
in higher dimensions. In addition, it might be possible to exploit the structure of a given 
independence problem to achieve better visualizations which is illustrated
for the conditional independence problem.

All associated significance tests are carried out using a conditional inference approach
instead of relying on unconditional asymptotic results.

>>>>

Relationships between categorical variables are typically
analyzed based on the underlying contingency tables that can be explored
for (in)dependence. Two standard methods from the statistical tool box
are log-linear models \citep[see, e.g.,][]{vcd:Agresti:2002} for modeling
independence and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
bringing it out graphically. Both methods can also be combined such that
a certain mosaic plot visualizes a particular log-linear model by 
controlling splitting order and direction and shading of the tiles
in the mosaic display
\citep{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
In particular, \cite{vcd:Friendly:1994} suggested a shading strategy
based on the residuals (typically, Pearson or deviance residuals) of the associated
log-linear model that elevates the mosaic plot from a display for frequencies
in a contingency table to a visualization technique that encompasses both, observed
frequencies and residuals. This shading allows for judging the quality of a model
fit and spotting dependence patterns that have not been accounted for by the model.
Other techniques for visualizing independence in contingency tables such as
association plots \citep{vcd:Cohen:1980} can also be enhanced by using this
residual-based shading.

In this paper, the shading of \cite{vcd:Friendly:1994} is extended in two directions:
usage of perceptually-based HCL (Hue-Chroma-Luminance) colors and combination of
visualization and significance
testing. Friendly's shading is typically implemented in statistical software using color
spaces whose dimensions are only poorly mapped to the perceptual dimensions of the
human visual system \citep{vcd:Brewer:1999,vcd:Ihaka:2003} which makes it more difficult
to properly read and interpret the corresponding plots. For overcoming this problem,
a general strategy for constructing diverging palettes in the perceptually-based HCL
space \citep{vcd:Ihaka:2003} is derived. To couple the visualization and the independence
model of a contingency table more tightly than in previous approaches, the residual-based
shading is extended such that presence of color in the display is equivalent to significance
of the associated independence test. This is achieved by using data-driven cut offs
for the appearance of color computed from the conditional permutation distribution
\citep{vcd:Ernst:2004,vcd:Pesarin:2001} of maximum-type test statistics. The resulting
residual-based shadings are illustrated both for mosaic and association plots 
using real world data sets.

The remainder of the paper is structured as follows: Section~\ref{sec:2way} gives a
brief introduction to significance tests and visualization techniques for the
independence problem in 2-way contingency tables. Based on this, Section~\ref{sec:shadings}
introduces the extended residual-based shadings using perceptually uniform HCL colors
and combining visualization and significance testing. The results are generalized to
multi-way tables in Section~\ref{sec:3way} with a more detailed discussion of adapting
them to conditional independence. Section~\ref{sec:conclusions} summarizes the paper
and gives some concluding remarks.

@@

\begin{table}[b!]
\begin{center}
\caption{\label{tab:art} Treatment and improvement among $59$ patients
with rheumatoid arthritis.}

@@

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced first in statistics
textbooks if they address the independence problem in 2-way tables
\citep[see, e.g.,][]{vcd:Agresti:2002}.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<classic_plots,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, margins = c(2.5, 4, 2.5, 3))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
assoc(art, newpage = FALSE, margins = c(5, 2, 5, 4))
popViewport(2)
@
\caption{Classic mosaic and association plot for the arthritis data.}
\label{fig:classic}
\end{center}
\end{figure}

@@

@@

\clearpage

\section{Extension to Conditional Independence} \label{sec:3way}

@@

\section{Conclusions} \label{sec:conclusions}

Various strategies for constructing residual-based shadings for visualizing
(conditional) independence in contingency tables via mosaic and association plots
are discussed. The shading
of \cite{vcd:Friendly:1994} is extended in two directions: the use of perceptually
uniform HCL colors and the combination of visualization and significance testing.
To achieve the former, a general guideline for constructing diverging palettes
in HCL space is introduced. The advantages of using this HCL shading scheme
instead of an HSV scheme are that the colors from this perceptually-based color space
provide uniform saturations over different hues and that the colorfulness in this
shading can be controlled independently from the other two dimensions.
%
To combine visualization and significance testing, two approaches are presented:
The first approach always uses a fully colored palette but relies on
data-driven cut offs such that the presence of color is equivalent to significance of the
associated maximum test. The second approach uses pre-defined cut offs (such as $2$ and $4$) but
codes the result of the significance test by using full colors only for significant results and
the same type of palette with a reduced amount of color for non-significant results.
%
Both results can not only be applied to the simple independence problem in 2-way
contingency tables, but also to arbitrary inpendence models fitted via log-linear models
in higher dimensions. In addition, it might be possible to exploit the structure of a given 
independence problem to achieve better visualizations which is illustrated
for the conditional independence problem.
%
All associated significance tests are carried out using a conditional inference approach
instead of relying on unconditional asymptotic results.
